---
title: "ACS data retrieval"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      message = FALSE,
                      results = 'hide',
                      warning = FALSE)
options(tinytex.verbose = TRUE)
```

A connection needs to be made in order to interface with the ACS APIs. Using the API key, which is unique for each user, the data can be converted into data frames. A tutorial of tidycensus can be found here: [Basic usage of tidycensus](https://walker-data.com/tidycensus/articles/basic-usage.html). Note that sometimes the connection can be lost if there are large pauses between steps and errors will appear, even if there is no actual error. Rerun prior processes just to be safe.  

```{r}
# use these packages

library(tidyverse)
library(tidycensus)
library(sf)
library(knitr)
```

1. Request the key to interface with ACS, which can be found on the [key signup](https://api.census.gov/data/key_signup.html) website. 
 
```{r}
# Once the key is saved, don't need to use it again for any following applications involving ACS APIs. 
#If you do, you need to overwrite, like the following.

census_api_key("5c21716c9afca860e0cbe0af36ce624318258e30",
               install = TRUE,
               overwrite = TRUE)
```

2. After interfacing,look at all variables in ACS-5yr survey to confirm the name of the variables of interest on the list. 

```{r}
# variable list from from the ACS-5yr (2015-2019) estimates
# use the filter to narrow down variables of interest. 
v19 <- load_variables(2019,
                      "acs5",
                      cache = TRUE)

view(v19)
```

3. Set up variables list. Go on the [census website](https://data.census.gov/cedsci/) if you want to confirm the correct names. Each table has a TableID that corresponds to the variable name. 

```{r}
# 4 indicators (income, race, employment, and age) and the total population variable
# since the tables did not have the exact values we were interested in, additional variables were selected to calculate them (ex: total population - population with income above 200% = population with income below 200%)
# total population variables represented by B01001_001: total population estimate from the 'sex by age' table

vars <- c(
          "C17002_008", ## RATIO OF INCOME TO POVERTY LEVEL IN THE PAST 12 MONTHS: #               Estimate!!Total:!!2.00 and over
          "C17002_001", # Estimate!!Total:
          "B03002_003", ## HISPANIC OR LATINO ORIGIN BY RACE: # Estimate!!Total:!!Hispanic or Latino:!!White alone
          "B03002_001", # Estimate!!Total:
          "B23025_007", ## EMPLOYMENT STATUS FOR THE POPULATION 16 YEARS AND OLDER: # Estimate!!Total:!!Not in labor force
          "B23025_001", # Estimate!!Total:
          "B01001_001", ## SEX BY AGE: # Estimate!!Total:
          "B01001_020", # Estimate!!Total:!!Male:!!65 and 66 years
          "B01001_021", # Estimate!!Total:!!Male:!!67 to 69 years
          "B01001_022", # Estimate!!Total:!!Male:!!70 to 74 years
          "B01001_023", # Estimate!!Total:!!Male:!!75 to 79 years
          "B01001_024", # Estimate!!Total:!!Male:!!80 to 84 years
          "B01001_025", # Estimate!!Total:!!Male:!!85 years and over
          "B01001_044", # Same as above, except female
          "B01001_045",
          "B01001_046",
          "B01001_047",
          "B01001_048",
          "B01001_049",
          "B01001_003", # Estimate!!Total:!!Male:!!Under 5 years
          "B01001_004", # Estimate!!Total:!!Male:!!5 to 9 years
          "B01001_005", # Estimate!!Total:!!Male:!!10 to 14 years
          "B01001_006", # Estimate!!Total:!!Male:!!15 to 17 years
          "B01001_027", # Same as avove, except female
          "B01001_028",
          "B01001_029",
          "B01001_030")
```

4. Connect to the ACS website with the necessary arguments. 

```{r} 
# note that for the resolution cannot get smaller than the county level, so narrowing down estimates for the city of Tacoma is reserved for the ArcGIS model

acs_connection <- get_acs(geography = "block group", # census block group
                              variables = vars, # chosen variable values
                              state = "WA", # Washington
                              county = "Pierce", # Pierce county
                              year = 2019, # ACS 5-yr (2015-2019) estimates
                              geometry = TRUE) # include polygons
```

5. Specify format of tibble for the final results. 

```{r}
acs_format <- acs_connection %>% 
   select(GEOID, NAME, variable, estimate, geometry) %>% # without selection, it will include measure of error, which is not necessary
                              spread(variable, estimate) %>% # formats a pair of key:value columns into a set of tidy columns 
                              rowwise()  # computes data a row at a time
```

6. Select the data tables of interest using the variables specified earlier. Transform through operations to get required final results.

```{r}
acs_vars <- acs_format %>% 
     mutate(pct_poverty = (C17002_001 - C17002_008) / C17002_001 * 100, # percent of population with ratio of income to poverty level in the past 2 months below 2.00
            pct_poc = (B03002_001 - B03002_003) / B03002_001 * 100, # percent of population with race other than white
            pct_unemployed = B23025_007 / B23025_001 * 100, # percent of population not in employment 16 years and older
            dep_ratio = sum(B01001_020, B01001_021, B01001_022, B01001_023, B01001_024, B01001_025, # dependency ratio: age-population ratio of the population above 65 and under 18 years older to the working-age population
                            B01001_044, B01001_045, B01001_046, B01001_047, B01001_048, B01001_049,
                            B01001_003, B01001_004, B01001_005, B01001_006, B01001_027, B01001_028,
                            B01001_029, B01001_030) / (B01001_001 - sum(B01001_020, B01001_021,
                                                                        B01001_022, B01001_023,
                                                                        B01001_024, B01001_025,
                                                                        B01001_044, B01001_045,
                                                                        B01001_046, B01001_047,
                                                                        B01001_048, B01001_049,
                                                                        B01001_003, B01001_004,
                                                                        B01001_005, B01001_006,
                                                                        B01001_027, B01001_028,
                                                                        B01001_029, B01001_030)) * 100,

            tot_pop = B01001_001) # total population
    
```

7. Set up format for final data table. 

```{r}
acs_final <- acs_vars %>% 
    select(GEOID,
           NAME,
           pct_poverty, # income indicator
           pct_poc, # race indicator
           pct_unemployed, # employment indicator
           dep_ratio, # age indicator
           tot_pop, # total population
           geometry)
```

8. Turn into shapefile, which can be imported into the Environmental Equity Index model.  

```{r}
# final check to make sure tibble fits expectations

head(acs_final) 

# naming the shapefile 

shp = "Pierce_county_5yr_acs_2019.shp"

st_write(acs_final, shp,
         append = FALSE)

```


